Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2020-05-05 paper.Rnw
took out overview (1.1)
given planned changes in response to #5 this subsection seemed redundant
also, luke didn't seem to like figure (#7)

added some TOC to the end of the intro (#1)

After a brief overview of the three causal approaches,
the next section will provide background for the case-study,
describing the CTA1 curriculum, the RAND study, and our focus on
hints.
The following three sections will present the observational study,
causal mediation, and principal stratification approaches to the hint
data.
As it turns out, these analyses appear to give contradictory
results---the observational study and mediation analysis estimate a
negative effect of requesting hints, while the principal
stratification analysis finds that students who request more hints
experience larger overall treatment effects.
Section \ref{sec:synthesis} will contrast these findings (including a
possible reconciliation), compare the three methods generally,
and discuss how they may be applied to data from other educational
technology RCTs.
Section \ref{sec:conclusion} will conclude.

@@

\subsection{Brief Overview of Three Causal Approaches}
\begin{figure}
\centering
\begin{tikzpicture}[scale=1,transform shape]
	%\draw[help lines] (0,0) grid (15,3);
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (1,0) {$H$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(0:3){$Y$};
	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
	%\draw[->,line width=2pt](x)--(z);
	%\draw[->,line width=2pt](x)--(y);
	\draw[->,line width=2pt](z)--(y);
	\node at (2.5,-1) {(a)};

	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (5,0) {$Z$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(8,0){$Y$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (m) at (6.5,2){$H$};
	\draw[->,line width=2pt](z)--(m);
	\draw[->,line width=2pt](m)--(y);
	\draw[->,line width=2pt](z)--(y);
	\node at (5.4,1.2) {(1)};
	\node at (7.6,1.2) {(2)};
	\node at (6.5,.3) {(3)};
	\node at (6.5,-1) {(b)};

        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (10,0) {$Z$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(12,0){$Y$};
        \node (mid) at (11,0) {};
        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (h)
        at (11,2) {$H(1)$};
	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
	%\draw[->,line width=2pt](x)--(z);
	%\draw[->,line width=2pt](x)--(y);
	\draw[->,line width=2pt](z)--(y);
        \draw[->,line width=2pt](h)--(mid);
	\node at (11,-1) {(c)};

\end{tikzpicture}
\caption{A schematic representation of three causal models for
  log data: (a) an observational study estimating the effects of hints
  $H$ on posttest scores $Y$, (b) a mediation analysis estimating
  natural direct effects of treatment assignment $Z$ on $Y$ and
  indirect effects mediated by $H$, and (c) principal stratification.}
\label{mediationFig}
\end{figure}

Figure \ref{mediationFig} is a schematic representation of the three
approaches to causal modeling of log data that we will discuss in this
paper.
First, an observational study to estimate the effect of hint usage $H$
on posttest scores $Y$, is represented by panel (a).
This approach, which we will discuss in Section
\ref{sec:observational}, discards the control group from the RCT and
uses techniques from observational studies to determine if and how requesting
a greater number of hints affects posttest scores.
For the sake of illustration, we will use techniques similar to those
discussed in \citet{rosenbaum2002observational},
 \citet{hansen2009b}, and \citet{ho:etal:2007}, combining a propensity
 score matching design with methods for analyzing experiments and
 parametric modeling.
The conceptual and methodological issues we will discuss extend to
other covariate control methods as well.

Next, as represented in panel (b), we will conduct a mediation
analysis, estimating natural
indirect effects of CTA1 treatment assignment $Z$, as mediated by
$H$. That is, we will estimate the component of $Z$'s effect on $Y$ that comes about
due to $Z$'s effect on $H$---path (1)---and $H$'s effect on $Y$, path
(2).
Notice that path (2) is identical to panel (a)---the effect of $H$ on
$Y$.
We will use this equivalence to estimate and interpret indirect
effects.
As part of the mediation analysis, we will also estimate path (3), the
natural direct effect of $Z$ on $Y$---that is, the component of $Z$'s
effect on $Y$ that is not due to hints $H$.

Finally, we will estimate a principal stratification model, as represented in panel
(c).
In the principal stratification analysis, we do not model $H$ as an
intervention that could, in theory, be randomized.
Instead, we posit a variable ``$H_i(1)$,''  subject $i$'s penchant for
requesting hints \emph{if} $i$ is assigned to the treatment
condition.
Unlike $H$, $H(1)$ is well defined (if
unobserved) for all participants in the RCT.
Principal stratification treats $H(1)$ as an effect moderator or
interaction variable: $H(1)$ does not affect $Y$, but the effect of
$Z$ on $Y$ may vary as a function of $H(1)$.
This analysis will largely follow the method of \citet{aoas}, which in
turn built heavily on \citet{schwartz2011bayesian} and
\citet{jin2008principal}.

All three of these approaches are designed to model intermediate
variables---those variables, like hint usage,
that were possibly affected by $Z$ but were measured prior to $Y$.
Though they were developed for more traditional data scenarios, the
novel scenario of computer log data from an educational technology RCT
presents new and interesting challenges and opportunities.

>>>>

The following section will provide background for the case-study,
describing the CTA1 curriculum, the RAND study, and our focus on
hints.
The following three sections will present the observational study,
which discards the control group altogether and estimates the effect
of hint requests on posttest scores;
causal mediation, which decomposes the overall treatment effect into a
component driven by hint requests and another driven by other
mechanisms; and principal stratification, which estimates separate
treatment effects for groups of students who would request hints
differently, if assigned to treatment.
As it turns out, these analyses appear to give contradictory
results---the observational study and mediation analysis estimate a
negative effect of requesting hints, while the principal
stratification analysis finds that students who request more hints
experience larger overall treatment effects.
Section \ref{sec:synthesis} will contrast these findings (including a
possible reconciliation), compare the three methods generally,
and discuss how they may be applied to data from other educational
technology RCTs.
Section \ref{sec:conclusion} will conclude.

@@
